Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data

### 2021-10-12 Manuscripts/Article/Article.tex
set up .tex manuscript file

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{url,hyperref}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,array,eucal, amsthm}
\linespread{1.5}
\setlength\parindent{35pt}
\input{macros}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{color}

\graphicspath{ {../Experiment Output/Figures/} }

\newcommand{\jx}[1]{{\color{blue}{ #1}}}

@@

	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
	A Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data
	}
	\author{Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
	\small $^{1}$Department of Statistical Science, Duke University \\
	}
	\date{\today}	
	\maketitle

	\begin{abstract}
		Stochastic epidemic models provide a realistic an interpretable description of the spread of a disease through a population. Yet, fitting these models in missing data settings is a notably difficult task; in particular, when the epidemic process is only partially observed, the likelihood of the model is intractable. To remedy this issue, this article introduces a data-augmented MCMC algorithm for fast and exact inference for the stochastic SIR model given discretely observed infection incidence counts. In a Metropolis-Hastings step, the algorithm jointly proposes event times of the augmented data according to a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data. Not only is the algorithm fast, but, since the augmented data are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the augmented data per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently. Moreover, this MCMC algorithm is to be uniformly ergodic. While existing MCMC approaches become intractable for populations greater than a few thousand individuals, unless they rely on model simplifications or approximations, the proposed algorithm enables exact inference and scales to outbreaks in populations of hundred-thousands individuals
		+ even on a single laptop. We validate its performance via thorough simulation experiments, and a case study on the $2014$ Ebola outbreak in Western Africa. % in Gu\'eck\'edou, Guinea. %, in a population of $150,000$ individuals. 
	\end{abstract}
	\keywords{Stochastic epidemic model; incidence data; exact posterior inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}

@@

	Conducting inference on SEMs is, however, a notably difficult task. The difficulty stems from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time. This makes the likelihood of the model intractable; were the infection and recovery times of each individual of the population observed, conducting inference on SEMs would be straightforward, but this is rarely the case. In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of people infectious at discrete times. The marginal likelihood of these partially observed data becomes a bottleneck, requiring a large integration step that accounts for all possible configurations of the missing data. In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form are available. Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs of their algorithm makes the approach applicable to small populations only.

	In this article, we introduce a data-augmented MCMC algorithm to estimate the parameters of the stochastic SIR model, a commonly used SEM, from discretely observed infection incidence counts. Rather than marginalizing directly, our approach accounts for the latent space via sampling. The algorithm updates the event times of the augmented data by jointly proposing infection and removal times from a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data. This surrogate process serves as an efficient proposal scheme within a Metropolis-Hastings algorithm. Since the latent spaces are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the event times per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently.

@@

	Our point of departure is to consider the stochastic SIR model -- also referred to as the general epidemic model -- which offers a parsimonious and interpretable representation of the mechanistic, population-level dynamics of an epidemic. The stochastic SIR model is a compartmental model in which the individuals of a population transition through three types or compartments: susceptible (S), infectious (I) and removed (R). The only possible moves are from S to I (infections) and from I to R (removals). A susceptible individual becomes infected through contact with an infectious individual. Once infected, she is immediately infectious and remains so for some period of time after which she is removed from the process without the possibility of reinfection. In this formulation, demographics dynamics such as births and deaths of individuals are ignored since they usually occur at a much slower rate than infections and removals.

	Assuming a closed population with a fixed size $n$, the stochastic SIR model consists of a continuous-time vector-valued process
	\begin{equation}
		\label{eq:X}
		\X = \left\lbrace \X(t), t>0\right\rbrace \in \chi_{\X}
	\end{equation}
	with
	\begin{equation}
		\X(t) = (X_1(t), \dots, X_n(t)) \in \{S, I, R\}^n
	\end{equation}
	where the agent-level subprocess
	$$ X_i(t) = 
	\begin{cases}
		S, & t \in [0, \tau^T_i] \\
		I, & t \in (\tau^T_i, \tau^J_i] \\
		R, & t \in (\tau^J_i, \infty)
	\end{cases}
	,\quad \iton
	$$
	denotes the status (compartment) of individual $i$ at time $t$ with $\tau^T_i$ and $\tau^J_i$ respectively the infection and removal times of individual $i$. If individual $i$ never becomes infectious, then we set $\tau^T_i = \tau^J_i = \infty$ and $X_i(t) = S$ for $t \in [0, \infty)$. The set $\chi_{\X}$ consists of the trajectories compatible with the evolution of a disease, that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted :
	\begin{equation}
		\label{eq:chi}
		\chi_{\X} = \{X:X(t) \in \{S,R\}^n \Rightarrow X(t+s) = X(t), \forall s>0 \}.
	\end{equation}

	The stochastic SIR model is specified by the rates at which individuals move from one compartment to another. If we assume a homogeneously mixing population where contacts between individuals occur independently at some rate constant $\beta$, then the contacts between two given individuals are said to follow a Poisson process with rate $\beta$. This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \ref{eq:X} is a time–homogeneous continuous time Markov chain whose instantaneous transition rates 
	%from state $\x$ to $\x'$
	are given by the matrix $\Lambda = [\lambda_{\x, \x'}]$ with
	$$
	\lambda_{\x, \x'} = 
	\begin{cases}
		\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
		\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
		0, & \text{ otherwise.}
	\end{cases}
	$$
	where $I(t) = \sum_{i=1}^n \mathbf{1} \{x_i(t) = I\}$ is the total number of infectious individuals at time $t$. That is, the individual-level infection and removal rate at time $t$ are respectively $\beta I(t)$ and $\gamma$.
	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar; also uncapitalize X on the right}
%	$$
%	P(\X_t = \x, \X_{t+dt} = \x') =
%	\begin{cases}
%		\beta I(t) dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
%		\gamma dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
%		o(dt), & \text{otherwise}
%	\end{cases}
%	$$
%	where $dt>0$ is small.

@@

	Assuming that the Markov process \ref{eq:X} is completely observed until time $t_{end}$, we obtain the following complete data likelihood
	\begin{align}
		L(\theta; \X)
		& = \prod_{T} \beta I(\tau^T_j) \prod_{J} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
		\label{eq:cdl}
		& = \beta^{n_T} \gamma^{n_J}\prod_{J} I(\tau^T_j) \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace
	\end{align}
	where $\theta = (\beta, \gamma) \in \chi_{\theta}$  are the parameters of the model and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ is the parameter space, 
	$T = \{i:\tau^T_i \le t_{end}\}$ and $J = \{i:\tau^J_i \le t_{end}\}$ are the respective index sets of individual infected and removed by time $t_{end}$, 
	$n_T = |T|$ and $n_J = |J|$ are the respective numbers of observed infections and removals,
	and	$S(t) = \sum_{i=1}^n \mathbf{1}\{X_i(t) = S\}$ is the number of susceptible individuals at time $t$.

	We see that in this fully observed scenario, it is straightforward to conduct inference. The likelihood \ref{eq:cdl} belongs to the exponential family and the maximum likelihood estimates can be expressed in terms of the sufficient statistics defined above as
	$$\hat{\beta}_{MLE} = \dfrac{n_T}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \quad \hat{\gamma}_{MLE} = \dfrac{n_J}{\int_{0}^{t_{end}} I(t)dt}.$$
	Since the functions $I$ and $S$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy of the gamma distribution: if we put independent gamma prior distributions on the parameters
		\begin{equation}
		\label{eq:pri}
		\beta \sim Ga(a_{\beta}, b_{\beta}), \qquad \gamma \sim Ga(a_{\gamma}, b_{\gamma})
	\end{equation}
	where $Ga(a,b)$ denotes the gamma distribution with mean $a/b$ and variance $a/b^2$, then the posterior distributions of the parameters are independent gamma distributions
	\begin{equation}
		\label{eq:pdb}
		\beta | \X \sim Ga\left( a_{\beta} + n_T, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right)
	\end{equation}
	and
	\begin{equation}
		\label{eq:pdg}
		\gamma | \X \sim Ga\left( a_{\gamma} + n_J, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
	\end{equation}
	from which we can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution $\pi(\theta|\X)$.

@@

	(New in this section: I am motivating the type of observed data  $T_{1:K}$ that we consider.)

	In practice, inference is complicated by the fact that the Markov process \ref{eq:X} is only partially observed. That is, we do not observe all epidemic events, and do not have access to the sufficient statistics needed to evaluate the likelihood \ref{eq:cdl}. Various types of partially observed data have been considered in the literature such as observing only the removal times \cite{Gibson.1998, ONeill.1999} and knowing the number of infectious individuals in the population at discrete points in time \cite{Fintzi.2017}. The former type of data for instance arise in animal experiment in which an animal that tests positive for the disease is immediately isolated from the rest of the population and therefore no longer contributes to the spread of the disease. In such case, the removal times are exactly known.

	In this article, we consider another type of partially observed data: incidence counts for the infections at discrete points in time. Such data arise when an infectious individual may continue to infect others even after being tested positive for the disease. This was for instance the case during the $2013$–$2016$ outbreak of Ebola virus. This pandemic was unprecedented in scale, with more than $10,000$ reported deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone). This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus. Numerous infections happened at the hospitals where infected individuals received a treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition. Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}. 
	It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.

	Consider an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$. The observed data consist of the $K$-dimensional vector $\Y = T_{1:K}$ where $T_k = \sum_{i=1}^n \mathbf{1}\{\tau^T_i \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ time interval.
	For simplicity, we further assume that the initial configuration of the population $\X(0)$ is known.

	If we adopt a Bayesian approach, the posterior distribution of the parameters given the observed data is
	\begin{align}
		\label{eq:pdl}
		\pi(\theta|\Y) 
		& \propto L(\theta; \Y)\pi(\theta) \nonumber \\
		& = \int_{\chi^*_{\X}} L(\theta; \x) \pi(\theta) d\x
	\end{align}
	where $\pi(\theta)$ is the prior distribution on the parameters $\theta$ and
	$\chi^*_{\X} \subset \chi_{\X}$ is the set of trajectories of the process \ref{eq:X} that are compatible with the observed data $\Y$.
	The partial data likelihood $L(\theta; \Y)$ therefore consists of a high-dimensional integral over all epidemic paths compatible with the observed data and has no closed form solution available, even for a population of moderate size.

	\subsection{Prior Work}
	\label{sec:pre}
	Researchers have explored several approaches for conducting inference on partially observed stochastic epidemic processes. Early approaches include the use of martingale-based equations \cite{Becker.1977, Watson.1981, Sudbury.1985}. However, such methods are difficult to apply to dynamic models with partially observed data and are therefore not suitable to the stochastic SIR model with incomplete data.
	More recently, researchers have based their computation on a simpler process that approximates the model's dynamics and whose likelihood in the presence of partially observed data is tractable. These approximation models include binomial chains models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}. While these approximation-based approaches bypass the intractability of the partial data likelihood \ref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.

@@

	%An advantage of DA-MCMC methods is the possibility to conduct inference in more complex models. For instance, Streftaris and Renshaw \cite{Streftaris.2002} generalized the approach of \cite{Gibson.1998} to a non-Markovian stochastic SIR model where the amount of time spent infectious is not necessarily exponentially distributed but follows an arbitrary distribution, and Bu et al. \cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network.

	\subsection{Proposed Data-Augmented Approach}
	\label{sec:con}

	TODO: Add short intro to not intimidating reader with notation.

	As shown in Section \ref{sec:icd}, the complete data likelihood \ref{eq:cdl} is amenable to computation. This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \ref{eq:cdl} and constructing a Markov chain on the space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}. We consider the latent data $\Z = \left\lbrace (z^T_i, z^J_i)\right\rbrace_{i=1}^{n} \in \chi_{\Z}$ which consist of the times and types of unobserved epidemic events, where the infection time of individual $i$ is
	$$z^T_i \begin{cases}
		\in [0, t_{end}) & \text{if individual } i \text{ becomes infectious before } t_{end} \\
		= \infty & \text{if individual } i \text{ does not become infectious before } t_{end} \\
	\end{cases}$$
	and the removal time of individual $i$ is
	$$z^J_i \begin{cases}
		\in (z_{T_i}, t_{end}] & \text{if individual } i \text{ is removed before } t_{end} \\
		= \infty & \text{if } z_{T_i} = \infty \text{ or individual } i \text{ is removed after } t_{end}. \\
	\end{cases}$$

@@

	% 1. convergence in medium-size population
	% 2. rho
	% 3. nominal coverage should be 95%.
	% 4. distribution of estimate for PD-SIR (unbiased) vs. LNA (biased)
	% explain why not compare chain binomial, particle filtering (forward simulation).

>>>>

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{url,hyperref}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,array,eucal, amsthm}
\linespread{1.5}
\setlength\parindent{35pt}
\input{../macros}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{color}

\graphicspath{ {../Figures/} }

\newcommand{\jx}[1]{{\color{blue}{ #1}}}
\newcommand{\ram}[1]{{\color{green}{ #1}}}

@@

	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
		Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data}
	\author{Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
		\small $^{1}$Department of Statistical Science, Duke University \\
	}
	\date{\today}	
	\maketitle

	\begin{abstract}
		Stochastic epidemic models provide an interpretable probabilistic description of the spread of a disease through a population. Yet, fitting these models in missing data settings is a notably difficult task; in particular, when the epidemic process is only partially observed, the likelihood of many classic models becomes intractable. To remedy this issue, this article introduces a data-augmented MCMC algorithm for fast and exact inference under the stochastic SIR model given discretely observed incidence counts. In a Metropolis-Hastings step, the algorithm jointly proposes event times of the augmented data according to a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data. Not only is the algorithm fast, but since the augmented data are generated from a very faithful approximation of the target epidemic model, the sampler can update large portions of the augmented data per iteration without lowering the acceptance rate prohibitively. We find that the method explores the high-dimensional latent space efficiently, and show that the Markov chain underlying the algorithm is uniformly ergodic. % While existing exact MCMC approaches become intractable for populations greater than a few thousand individuals,
		The proposed algorithm enables exact inference and scales to outbreaks in populations of hundreds of thousands, even on a single laptop. We validate its performance via thorough simulation experiments, consider and a case study on the $2014$ Ebola outbreak in Western Africa. % in Gu\'eck\'edou, Guinea. %, in a population of $150,000$ individuals. test. 
	\end{abstract}
	\keywords{Compartmental model; incidence data; exact posterior inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}

@@

	Conducting inference on SEMs is, however, a notably difficult task. Challenges stem from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time, making the likelihood of the model intractable. Were the infection and recovery times of each individual of the population observed, conducting inference on SEMs would be straightforward, but this is rarely the case. In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of people infectious at discrete reporting times. For instance, the $2013$–$2016$ outbreak of Ebola virus reported over $10,000$, mainly in three Western African countries (Guinea, Liberia and Sierra Leone). This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data. Numerous infections happened at the hospitals where infected individuals received a treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition. Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}. 
	It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time \jx{to do: stronger transition tying back to goal of partial observed inference}.

	More specifically, the marginal likelihood of such partially observed data becomes a bottleneck, as it requires a large integration step that accounts for all possible configurations of the missing. In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form is available. \jx{to do: cite recent papers from Simon Spencer's group too that do so for discrete-time models, and mention classical matrix exponentiation being intractable}. Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs limit their algorithm to moderate sized outbreaks.

	In this article, we make use of data augmentation to explore configurations of the missing data through latent variables.  We show how a data-augmented MCMC algorithm can target the exact joint posterior of model parameters and these latent variables under the stochastic SIR model, a commonly used SEM, from discretely observed infection incidence counts. Rather than marginalizing directly, our approach accounts for the latent space via sampling. The algorithm updates the event times of the augmented data by jointly proposing infection and removal times from a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data. Its success lies in the design of this surrogate process serving as an efficient proposal scheme within a Metropolis-Hastings algorithm. Since the latent spaces are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the event times per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently.

@@

	Our point of departure is to consider the stochastic SIR model -- also referred to as the general epidemic model -- which offers a parsimonious and interpretable representation of the mechanistic, population-level dynamics of an epidemic. The stochastic SIR model is a compartmental model in which the individuals of a population transition through three types or compartments: susceptible (S), infectious (I) and removed (R). The only possible moves are from S to I (infections) and from I to R (removals). A susceptible individual becomes infected through contact with an infectious individual. Once infected, she is immediately infectious and remains so for some period of time after which she is removed from the process without the possibility of reinfection. In this formulation, demographic dynamics such as births and deaths of individuals are ignored since they usually occur at a much slower rate than infections and removals.

	Assuming a closed population with a fixed size $n$, the stochastic SIR model consists of a continuous-time vector-valued process
	\begin{equation}
		\label{eq:X}
		\X = \left\lbrace \X(t), t>0\right\rbrace \in \chi_{\X}
	\end{equation}
	with
	\begin{equation}
		\X(t) = (X_1(t), \dots, X_n(t)) \in \{s, i, r\}^n
	\end{equation}
	where the agent-level subprocess
	$$ X_j(t) = 
	\begin{cases}
		s, & t \in [0, \tau^I_j] \\
		i, & t \in (\tau^I_j, \tau^R_j] \\
		r, & t \in (\tau^R_j, \infty)
	\end{cases}
	,\quad \iton
	$$
	denotes the status (compartment) of individual $j$ at time $t$ with $\tau^I_j$ and $\tau^R_j$ respectively the infection and removal times of individual $j$. If individual $j$ never becomes infectious, then we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$. Here $\chi_{\X}$ denotes the set of valid trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted :
	\begin{equation}
		\label{eq:chi}
		\chi_{\X} = \{X:X(t) \in \{s,r\}^n \Rightarrow X(t+u) = X(t), \forall u>0 \}.
	\end{equation}

	The stochastic SIR model is specified by the rates at which individuals move from one compartment to another. If we assume a homogeneously mixing population where contacts between individuals occur independently at some rate constant $\beta$, then the contacts between two given individuals are said to follow a Poisson process with rate $\beta$. This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time–homogeneous continuous time Markov chain, whose instantaneous transition rates 
	%from state $\x$ to $\x'$
	are given by the matrix $\Lambda = [\lambda_{\x, \x'}]$ with
	$$
	\lambda_{\x, \x'} = 
	\begin{cases}
		\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = s \text{ and } x_j' = i, \\
		\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = i \text{ and } x_j' = r, \\
		0, & \text{ otherwise.}
	\end{cases}
	$$
	where $I(t) = \#\{x_j(t) = i\}$ is the total number of infectious individuals at time $t$. Thus, the individual-level infection and removal rates at time $t$ are respectively $\beta I(t)$ and $\gamma$.
	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar; also uncapitalize X on the right}
	%	$$
	%	P(\X_t = \x, \X_{t+dt} = \x') =
	%	\begin{cases}
	%		\beta I(t) dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
	%		\gamma dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
	%		o(dt), & \text{otherwise}
	%	\end{cases}
	%	$$
	%	where $dt>0$ is small.

@@

	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood
	\begin{align}
		L(\theta; \X)
		& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
		\label{eq:cdl}
		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace
	\end{align}
	describing the complete data trajectory.
	To establish notation,
	$\theta = (\beta, \gamma) \in \chi_{\theta}$ are the model parameters 
	and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ denotes the parameter space, 
	the index sets $\mathcal{I} = \{j\in (1, \cdots, n): \tau^I_j \in (0, t_{end}]\}$ and $\mathcal{R} = \{j\in (1, \cdots, n): \tau^R_j \in (0, t_{end}]\}$ denote the individuals that are respectively infected and removed during the observation interval $(0, t_{end}]$,
	$n_I = |\mathcal{I}|$ and $n_R = |\mathcal{R}|$ are the numbers of observed infections and removals,
	and	$S(t) = \#\{X_i(t) = S\}$ is the number of susceptible individuals at time $t$.

	It is straightforward to conduct inference in this continuously observed scenario. In particular, the likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as
	$$\hat{\beta}_{MLE} = \dfrac{n_I}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \quad \hat{\gamma}_{MLE} = \dfrac{n_R}{\int_{0}^{t_{end}} I(t)dt}.$$
	Since the functions $I$ and $S$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy with the Gamma distribution: if we let the parameters follow independent Gamma prior distributions
	\begin{equation}
		\label{eq:pri}
		\beta \sim Ga(a_{\beta}, b_{\beta}), \qquad \gamma \sim Ga(a_{\gamma}, b_{\gamma})
	\end{equation}
	where $Ga(a,b)$ denotes the parametrization with mean $a/b$ and variance $a/b^2$, then the posterior distributions of the parameters are independent Gamma distributions
	\begin{equation}
		\label{eq:pdb}
		\beta | \X \sim Ga\left( a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right)
	\end{equation}
	and
	\begin{equation}
		\label{eq:pdg}
		\gamma | \X \sim Ga\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
	\end{equation}
	from which one can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution of parameters $\pi(\theta|\X)$.

@@

	%	(New in this section: I am motivating the type of observed data  $T_{1:K}$ that we consider.)

	In practice, inference is complicated by the fact that the process \eqref{eq:X} is only partially observed. When we do not observe all epidemic events, which is virtually always the case with real data, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}. Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals in the population at discrete points in time \cite{Fintzi.2017}. For instance, the former arise in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread. In such cases, exact times of removals are known, but such information is unavailable in typical observational studies.

	In this article, we focus on partially observed \textit{incidence} counts of infections at discrete points in time. Such data arise when an infectious individual may be identified some time after onset, and may continue to infect others even after being tested positive for the disease. %This was for instance the case during the $2013$–$2016$ outbreak of Ebola virus. This pandemic was unprecedented in scale, with more than $10,000$ reported deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone). This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus. Numerous infections happened at the hospitals where infected individuals received a treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition. Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}. 
	%It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.
	Consider an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$. The observed data consist of the $K$-dimensional vector $\Y = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ time interval.
	For simplicity, we further assume that the initial configuration of the population $\X(0)$ is known.

	If we adopt a Bayesian approach, the posterior distribution of the parameters given the observed data is formally related to the joint posterior via integration:
	\begin{equation}
		\label{eq:pdl}
		\pi(\theta|\Y) 
		\, \, \propto \, \, L(\theta; \Y)\pi(\theta) \, \, = \, \, \int_{\chi^*_{\X}} L(\theta; \x) \pi(\theta) d\x
	\end{equation}
	where $\pi(\theta)$ is the prior distribution on the parameters $\theta$ and
	$\chi^*_{\X} \subset \chi_{\X}$ is the set of trajectories of the process \eqref{eq:X} that are compatible with the observed data $\Y$.
	The partial data likelihood $L(\theta; \Y)$ therefore consists of a high-dimensional integral over all epidemic paths compatible with the observed data. This marginalization step has no closed form solution, and presents computational challenges even for a population of moderate size.

	\subsection{Prior Work}
	\label{sec:pre}
	Researchers have explored several approaches for conducting inference on partially observed stochastic epidemic processes. Early approaches include the use of martingale-based equations \cite{Becker.1977, Watson.1981, Sudbury.1985}. However, such methods are difficult to apply to dynamic models with partially observed data and are therefore not suitable to the stochastic SIR model with incomplete data.
	More recently, researchers have based their computation on a simpler process that approximates the model's dynamics and whose likelihood in the presence of partially observed data is tractable. Popular approximations include chain binomial models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}. While these approximation-based approaches bypass the intractability of the partial data likelihood \ref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.

@@

	%An advantage of DA-MCMC methods is the possibility to conduct inference in more complex models. For instance, Streftaris and Renshaw \cite{Streftaris.2002} generalized the approach of \cite{Gibson.1998} to a non-Markovian stochastic SIR model where the amount of time spent infectious is not necessarily exponentially distributed but follows an arbitrary distribution, and Bu et al. \cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network, and Kypraios (2018, 2020) model the infection dynamics nonparameterically.

	\section{Exact Inference with a  Data-Augmented Approach}
	\label{sec:con}

	We adopt a data-augmented MCMC approach that bridges the challenging partially observed setting to the tractable complete data likelihood by way of latent variables. In particular, our method hinges on the efficacy of a carefully designed proposal process for the latent variables that faithfully approximates the original process dynamics. 
	%By correcting with a Metropolis-Hastings step
	The sampler targets the exact posterior of the model given partially observed incidence data and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.

	As shown in Section \ref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation. This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl} and constructing a Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ with state space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}. We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n} \in \chi_{\Z}$ which consist of the times and types of unobserved epidemic events, where the infection and removal times for individual $j$ are
	$$z^I_j \begin{cases}
		\in [0, t_{end}) & \text{if individual } j \text{ becomes infectious before } t_{end} \\
		= \infty & \text{if individual } j \text{ does not become infectious before } t_{end} \\
	\end{cases}$$
	and
	$$z^R_j \begin{cases}
		\in (z^I_j, t_{end}] & \text{if individual } j \text{ is removed before } t_{end} \\
		= \infty & \text{if } z^I_j = \infty \text{ or individual } j \text{ is removed after } t_{end}. \\
	\end{cases}$$

@@

	% 1. convergence in medium-size population
	% 2. rho
	% 3. nominal coverage should be 95%: boxplot of posterior means, including coverage (one setting is enough for the moment)
	% 4. distribution of estimate for PD-SIR (unbiased) vs. LNA (biased)
	% explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data.

	% TODO: Tables: posterior means, credible intervals, effective sample size.
	% TODO: frequentist coverage for 3 different observation schedules
